---
title: "PA1_template"
author: "Per Rynning"
date: "7. januar 2016"
output: html_document
---

This document details my submission for course project 1 in the Reproducible Research course. 

The goal of the project is to download and analyze Acivity monitoring data and answer the following main questions: 

- What is mean total number of steps taken per day?
- What is the average daily activity pattern?
- Are there differences in activity patterns between weekdays and weekends?

To start off we will load the required libraries:

```{r, echo=TRUE, warning=FALSE, message=FALSE}
library(lubridate)
library(stringr)
library(dplyr)
library(ggplot2)

```

Next step is to download the data from <https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip> unless it is already present in the current working directory.

```{r}

## Downloads dataset if not available in the current folder
if(!file.exists("repdata-data-activity.zip") && !file.exists("activity.csv")){
    download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", 
                  destfile = "repdata-data-activity.zip")
    
    dateDownloaded <- date()
    
    ##This report was based on data downloaded `r dateDownloaded`.
    
    unzip("repdata-data-activity.zip")
}

```



We read the data using *read.csv* 

```{r}
## Reads the data from file
activity <- read.csv(file = "activity.csv", 
                     header = T)

```

We create a new column, *datetime*, by combining the *date* and *interval* fields, and convert this so that we have a POSIX date for each entry. 

```{r}
## Creates datetime column combining date and interval to POSIX date
activity$datetime <- paste(activity$date, 
                           str_pad(activity$interval, 
                                   width = 4, 
                                   side="left", 
                                   pad="0"))

activity$datetime <- ymd_hm(activity$datetime)
```

## Mean total number of steps

To make a histogram of total number of steps pr day we create a new data frame called *steps_pr_day*
```{r}
## Finds nr of steps pr day

steps_pr_day <- group_by(activity, 
                         date) %>% 
    summarise(totalsteps = sum(steps))

```

We then get this histogram showing total number of steps pr day  and the frequency.
**Note**: For this part of the exercise we have ignored missing values in the dataset. 
```{r}
hist(steps_pr_day$totalsteps, 
     main = "Frequency of total steps for a day", 
     xlab = "Total steps for a day")

```

The mean and median steps per day are

```{r}
## Mean steps per day
mean(steps_pr_day$totalsteps, na.rm = T)

## Median steps per day
median(steps_pr_day$totalsteps, na.rm = T)
```

## Average daily activity pattern

We'd like to see how the average daily activity pattern is. We can do this by grouping the data on the intervals.

```{r}
steps_pr_interval <- group_by(activity, 
                              interval) %>% 
    summarise(meansteps = mean(steps, 
                               na.rm = T))

## Uses qplot to plot the average daily activity pattern
qplot(interval, 
      meansteps, 
      data = steps_pr_interval, 
      type="line", 
      ylab = "Mean number of steps", 
      xlab = "Interval", 
      main = "Average daily activity pattern")

```

### Missing values
The activity dataset we are working with currently has many missing values

```{r}
## Sums up missing values from activity$steps
sum(is.na(activity$steps))
```

